---
title: "Alberta R-Value"
author: "Tony Nickonchuk"
date: "12/3/2020"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#load libraries
library(dplyr)
library(ggplot2)
library(EpiEstim)
library(incidence)
library(lubridate)
```

## What Is the R-Value?

The R-Value, or R0, or R-naught (take your pick), in simple terms, is the average number of people an infected person will go on to infect. It is not a perfect parameter, but generally, if R0 is below 1, an infection will die out in a population. If it is above 1, it will continue spreading.

That is why when Premier Jason Kenney of Alberta announced new public health restrictions, he said that the province was aiming for the R0 to be below 1 before they considered lifting the restrictions.

Initially, the government said they would be posting the R0 publicly as of November 30th, but [today backtracked](https://cisnfm.com/news/7489224/alberta-covid-19-r-value/) and said they will no longer be doing that. They are blaming the confusion on "human error". Because of course they are.

Well, if they aren't going to publish the R0, I will. Now, I don't have access to the internal R0 being used by Alberta Health. But I do know how to run a program to calculate an estimate of it from publicly available data. So I will do that below. I likely will only be able to update it weekly, as I'm doing this on my own time, of which I have very little.

Full code for this is available at my [GitHub repository](https://github.com/tonynick/abrvalue). Conversely, you can just take in the information.

```{r,include=F}
#download COVID-19 case data from AB Health
#go here https://www.alberta.ca/stats/covid-19-alberta-statistics.htm
#then click on "Data Export";scroll to the bottom and download
df<-read.csv("~/Downloads/covid19dataexport.csv",stringsAsFactors = F)
df$date<-ymd(df$Date.reported)
df$zone<-df$Alberta.Health.Services.Zone
df<-df%>%select(date,zone)
cal<-df%>%filter(zone=="Calgary Zone")
edm<-df%>%filter(zone=="Edmonton Zone")
sou<-df%>%filter(zone=="South Zone")
nor<-df%>%filter(zone=="North Zone")
cen<-df%>%filter(zone=="Central Zone")
ab<-df
xcal<-df%>%filter(zone!="Calgary Zone")

ab<-incidence(ab$date)
cal<-incidence(cal$date)
edm<-incidence(edm$date)
sou<-incidence(sou$date)
nor<-incidence(nor$date)
cen<-incidence(cen$date)
xcal<-incidence(xcal$date)

#serial interval mean and standard deviation taken from meta-analysis of 
#COVID serial interval here; used random effects estimate #https://www.ncbi.nlm.nih.gov/pmc/articles/PMC7448781/
#posterior CV default 0.3 for all; south and central could not reach that
#due to low incidence initially; increased by 0.1 until reached

abr<-estimate_R(ab,method="parametric_si",config=make_config(list(mean_si=5.19,std_si=0.418)))
calr<-estimate_R(cal,method="parametric_si",config=make_config(list(mean_si=5.19,std_si=0.418)))
edmr<-estimate_R(edm,method="parametric_si",config=make_config(list(mean_si=5.19,std_si=0.418)))
sour<-estimate_R(sou,method="parametric_si",config=make_config(list(mean_si=5.19,std_si=0.418,cv_posterior=0.4)))
norr<-estimate_R(nor,method="parametric_si",config=make_config(list(mean_si=5.19,std_si=0.418)))
cenr<-estimate_R(cen,method="parametric_si",config=make_config(list(mean_si=5.19,std_si=0.418,cv_posterior=0.6)))
xcalr<-estimate_R(xcal,method="parametric_si",config=make_config(list(mean_si=5.19,std_si=0.418)))

#remove first 7 days as r0 is calculated in rolling 7 day windows so first 7
#days are NA
ab<-data.frame(date=abr$dates[-c(1:7)],abr$R)
cal<-data.frame(date=calr$dates[-c(1:7)],calr$R)
edm<-data.frame(date=edmr$dates[-c(1:7)],edmr$R)
sou<-data.frame(date=sour$dates[-c(1:7)],sour$R)
nor<-data.frame(date=norr$dates[-c(1:7)],norr$R)
cen<-data.frame(date=cenr$dates[-c(1:7)],cenr$R)
xcal<-data.frame(date=xcalr$dates[-c(1:7)],xcalr$R)

#set to last 30 days
today<-ymd(Sys.Date())
cutoff<-today-31
ab<-ab%>%filter(date>=cutoff)
cal<-cal%>%filter(date>=cutoff)
edm<-edm%>%filter(date>=cutoff)
sou<-sou%>%filter(date>=cutoff)
nor<-nor%>%filter(date>=cutoff)
cen<-cen%>%filter(date>=cutoff)
xcal<-xcal%>%filter(date>=cutoff)

ab$time<-1:30
cal$time<-1:30
edm$time<-1:30
sou$time<-1:30
cen$time<-1:30
nor$time<-1:30
xcal$time<-1:30

seq<-seq(1,30,by=7)
ab$date_lab<-paste(month.abb[month(ab$date)],day(ab$date))
cal$date_lab<-ab$date_lab
edm$date_lab<-ab$date_lab
sou$date_lab<-ab$date_lab
nor$date_lab<-ab$date_lab
cen$date_lab<-ab$date_lab
xcal$date_lab<-ab$date_lab

ab<-ab%>%select(date,date_lab,time,Mean.R.,Quantile.0.025.R.,Quantile.0.975.R.)
names(ab)<-c("date","date_lab","time","mean","lower","upper")
cal<-cal%>%select(date,date_lab,time,Mean.R.,Quantile.0.025.R.,Quantile.0.975.R.)
names(cal)<-c("date","date_lab","time","mean","lower","upper")
edm<-edm%>%select(date,date_lab,time,Mean.R.,Quantile.0.025.R.,Quantile.0.975.R.)
names(edm)<-c("date","date_lab","time","mean","lower","upper")
sou<-sou%>%select(date,date_lab,time,Mean.R.,Quantile.0.025.R.,Quantile.0.975.R.)
names(sou)<-c("date","date_lab","time","mean","lower","upper")
nor<-nor%>%select(date,date_lab,time,Mean.R.,Quantile.0.025.R.,Quantile.0.975.R.)
names(nor)<-c("date","date_lab","time","mean","lower","upper")
cen<-cen%>%select(date,date_lab,time,Mean.R.,Quantile.0.025.R.,Quantile.0.975.R.)
names(cen)<-c("date","date_lab","time","mean","lower","upper")
xcal<-xcal%>%select(date,date_lab,time,Mean.R.,Quantile.0.025.R.,Quantile.0.975.R.)
names(xcal)<-c("date","date_lab","time","mean","lower","upper")
```

## Plots

All the plots are on the same y-axis scale of 0.5 to 2.0.

```{r alberta, echo=FALSE}
ggplot(ab,aes(x=time,y=mean))+geom_line()+geom_ribbon(aes(ymin=lower,ymax=upper),alpha=0.3)+ylim(c(0.5,2))+geom_vline(xintercept=ab$time[ab$date=="2020-11-24"],col="red",lty="dashed")+geom_hline(yintercept=1,col="black",lty="dashed")+annotate("text",y=1,x=1,hjust=0,vjust=-0.5,label="Target R-Value")+annotate("text",y=1,x=ab$time[ab$date=="2020-11-24"],hjust=1,vjust=-0.5,label="New Measures",angle=90)+labs(x="Date",y="R-Value")+ggtitle("R-Value Over Last 30 Days in Alberta")+scale_x_continuous(breaks=seq,labels=ab$date_lab[seq])
```

```{r calgary, echo=FALSE}
ggplot(cal,aes(x=time,y=mean))+geom_line()+geom_ribbon(aes(ymin=lower,ymax=upper),alpha=0.3)+ylim(c(0.5,2))+geom_vline(xintercept=cal$time[cal$date=="2020-11-24"],col="red",lty="dashed")+geom_hline(yintercept=1,col="black",lty="dashed")+annotate("text",y=1,x=1,hjust=0,vjust=-0.5,label="Target R-Value")+annotate("text",y=0.99,x=cal$time[cal$date=="2020-11-24"],hjust=1,vjust=-0.5,label="New Measures",angle=90)+labs(x="Date",y="R-Value")+ggtitle("R-Value Over Last 30 Days in Calgary Zone")+scale_x_continuous(breaks=seq,labels=cal$date_lab[seq])
```

```{r edmonton, echo=FALSE}
ggplot(edm,aes(x=time,y=mean))+geom_line()+geom_ribbon(aes(ymin=lower,ymax=upper),alpha=0.3)+ylim(c(0.5,2))+geom_vline(xintercept=cal$time[cal$date=="2020-11-24"],col="red",lty="dashed")+geom_hline(yintercept=1,col="black",lty="dashed")+annotate("text",y=1,x=1,hjust=-0.2,vjust=-0.5,label="Target R-Value")+annotate("text",y=0.99,x=cal$time[cal$date=="2020-11-24"],hjust=1,vjust=-0.5,label="New Measures",angle=90)+labs(x="Date",y="R-Value")+ggtitle("R-Value Over Last 30 Days in Edmonton Zone")+scale_x_continuous(breaks=seq,labels=cal$date_lab[seq])
```

```{r south, echo=FALSE}
ggplot(sou,aes(x=time,y=mean))+geom_line()+geom_ribbon(aes(ymin=lower,ymax=upper),alpha=0.3)+ylim(c(0.5,2))+geom_vline(xintercept=cal$time[cal$date=="2020-11-24"],col="red",lty="dashed")+geom_hline(yintercept=1,col="black",lty="dashed")+annotate("text",y=1,x=1,hjust=0,vjust=-0.5,label="Target R-Value")+annotate("text",y=.99,x=cal$time[cal$date=="2020-11-24"],hjust=1,vjust=-0.5,label="New Measures",angle=90)+labs(x="Date",y="R-Value")+ggtitle("R-Value Over Last 30 Days in South Zone")+scale_x_continuous(breaks=seq,labels=cal$date_lab[seq])
```

```{r north, echo=FALSE}
ggplot(nor,aes(x=time,y=mean))+geom_line()+geom_ribbon(aes(ymin=lower,ymax=upper),alpha=0.3)+ylim(c(0.5,2))+geom_vline(xintercept=cal$time[cal$date=="2020-11-24"],col="red",lty="dashed")+geom_hline(yintercept=1,col="black",lty="dashed")+annotate("text",y=1,x=1,hjust=0,vjust=-0.5,label="Target R-Value")+annotate("text",y=.99,x=cal$time[cal$date=="2020-11-24"],hjust=1,vjust=-0.5,label="New Measures",angle=90)+labs(x="Date",y="R-Value")+ggtitle("R-Value Over Last 30 Days in North Zone")+scale_x_continuous(breaks=seq,labels=cal$date_lab[seq])
```

```{r central, echo=FALSE}
ggplot(cen,aes(x=time,y=mean))+geom_line()+geom_ribbon(aes(ymin=lower,ymax=upper),alpha=0.3)+ylim(c(0.5,2))+geom_vline(xintercept=cal$time[cal$date=="2020-11-24"],col="red",lty="dashed")+geom_hline(yintercept=1,col="black",lty="dashed")+annotate("text",y=1,x=1,hjust=0,vjust=-0.5,label="Target R-Value")+annotate("text",y=0.99,x=cal$time[cal$date=="2020-11-24"],hjust=1,vjust=-0.5,label="New Measures",angle=90)+labs(x="Date",y="R-Value")+ggtitle("R-Value Over Last 30 Days in Central Zone")+scale_x_continuous(breaks=seq,labels=cal$date_lab[seq])
```

```{r ab exc cal, echo=FALSE}
ggplot(xcal,aes(x=time,y=mean))+geom_line()+geom_ribbon(aes(ymin=lower,ymax=upper),alpha=0.3)+ylim(c(0.5,2))+geom_vline(xintercept=cal$time[cal$date=="2020-11-24"],col="red",lty="dashed")+geom_hline(yintercept=1,col="black",lty="dashed")+annotate("text",y=1,x=1,hjust=0,vjust=-0.5,label="Target R-Value")+annotate("text",y=0.99,x=cal$time[cal$date=="2020-11-24"],hjust=1,vjust=-0.5,label="New Measures",angle=90)+labs(x="Date",y="R-Value")+ggtitle("R-Value Over Last 30 Days in Alberta Minus Calgary Zone")+scale_x_continuous(breaks=seq,labels=cal$date_lab[seq])
```